Supporting Information to Estimating the Initial Susceptible Fraction in Dengue Dynamics"

### 2014-10-13 S1_lm1.tex
moving Rt estimation to main text

\documentclass[12pt]{article}              
\usepackage[margin=1.5cm]{geometry}
\usepackage{setspace,relsize}               
\usepackage{moreverb}                        
\usepackage{url}
\usepackage{hyperref}
\hypersetup{colorlinks=true,citecolor=blue}
\usepackage{amsmath}
\usepackage{mathtools} 
\usepackage{amssymb}
\usepackage{indentfirst}
\usepackage[authoryear,round]{natbib}
\bibliographystyle{apalike}
\usepackage[pdftex]{lscape}
\usepackage[toc,page]{appendix}
% \usepackage{float}
% \usepackage{longtable}
% \usetikzlibrary{arrows}
%%%%
\title{
Supporting Information to ¨Estimating the Initial Susceptible Fraction in Dengue Dynamics"
}           
\author{Flavio Code\c{c}o Coelho \\
\\
\& \\
Luiz Max de Carvalho \\
}

\date{\today}
% Notation defs
\def \rr {$R_{t}\ $}

@@

In monitoring of infectious diseases, it is important to assess whether the incidence of a a particular disease is increasing significantly, in order to decide to take preventive measures.
The effective reproductive number at time $t$, \rr, can be understood as a real-time estimate of the basic reproductive number ($R_{0}$) and is defined as the average number of secondary cases per primary case at time $t$.
Here we derive confidence/credibility bounds for the uncertainty about the estimate of the effective reproductive number $R_t$ under the assumption that reported disease counts are Poisson distributed.
It is shown that one can construct a conservative confidence interval for \rr using the induced probability distribution that results from the approach of \citet{mantel}.

Let $Y_t$ be the number of reported disease cases for a particular time $t \in (0, T)$.
\citet{nishiura} extend the theory developed by \citet{stallybrass} and propose to estimate \rr as
\begin{equation}
\label{eq:Rtestimate}
R_t = \left( \frac{Y_{t+1}}{Y_t}\right)^{1/n}
\end{equation}
where $n$ is taken to be the ratio between the length of reporting interval and the mean generation time of the disease.
Here we are interested in the simpler case $n=1$.
If \rr is to be used as a decision tool, one needs to be able to quantify the uncertainty about estimate in equation~\ref{eq:Rtestimate}. 
Here we detail how to obtain confidence intervals for \rr, exploring two approaches under the assumption that the counts $Y_t$ are Poisson distributed for all $t$.

\section*{The approach Ederer \& Mantel (1974)}

\label{sec:mantel}
First we explore the approach of \citet{mantel}, whose objective is to obtain confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ and define $S = Y_{t} + Y_{t+1}$.
The authors then proceed to note that by conditioning on the sum $S$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(S, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that $Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that $Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
\citet{mantel} show that one can construct a $100\alpha \%$ confidence interval for \rr by noting that
\begin{align}
 R_t^{(L)} &= \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})}\\
 R_t^{(U)} &= \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}
\end{align}

Many authors have derived $c_{\alpha}(\theta_t)$ following orthodox approaches such as those of \citet{wilson} and \citet{clopper}, mainly for simplicity reasons.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the binomial likelihood in (\ref{eq:binlike}), the posterior distribution for $\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, S) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Equation~(\ref{eq:thetapost}) tells us that the induced distribution of $R_t$ is a beta prime with parameters $ a_1 = Y_{t+1} + a_0$ and $b_1 =  Y_t + b_0$.
The density of this induced distribution  $f_P(R_t)$ is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 - 1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr under $f_P$ is $a_1/(b_1 - 1)$ and its variance is $a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.

Since when $R_t > 1$ indicates sustained transmission, one may be interested in computing the probability of this event.
By noting that
\begin{align}
\label{cumprobMantel}
Pr(R_t > 1) &= 1 - \int_0^1 f_P(r)dr \\
            &= 1- Pr(\theta_t < \frac{1}{2})
\end{align}
one can avoid dealing with the density in~(\ref{eq:densityMantel}) directly.

\subsection*{A remark on prior distributions and tail behaviour}
\label{sec:tails}

In order to decide which approach to take, it may be of use analysing the variance and tail behaviour of the derived distributions for \rr. 
Consider the case of using a flat $Uniform(0, 1)$ prior for $\theta_t$ in Section~\ref{sec:mantel}.
With $a_0 = b_0 = 1$, $a_1 = a_2$ and $b_1 = b_2 + 1$.
The beta prime (inverse beta distribution) will have heavier tails compared to the conditional distribution in (\ref{eq:densityNishiura}), thus providing more conservative confidence/credibility intervals.  
As a side note, the Bayesian approach presented in Section~\ref{sec:mantel} will give similar results to those of \citet{wilson} and \citet{wilson} for $Y_{t+1}$ and $Y_t >> 1$.
Under the flat  uniform prior for $\theta_t$, the Bayesian posterior credibility interval is nearly indistinguishable from the confidence interval proposed by \citet{clopper} for $Y_{t+1}, Y_t > 20$.
Note that the $Beta(1, 1)$ uniform prior for $\theta_t$ constitutes a poor prior choice mainly because the induced distribution for \rr is only well-defined for $b_0 > 2$.

>>>>

\documentclass[12pt]{article}              
\usepackage[margin=2.0cm]{geometry}
\usepackage{setspace,relsize}               
\usepackage{moreverb}                        
\usepackage{url}
\usepackage{hyperref}
\hypersetup{colorlinks=true,citecolor=blue}
\usepackage{amsmath}
\usepackage{mathtools} 
\usepackage{amssymb}
\usepackage{indentfirst}
\usepackage[authoryear,round]{natbib}
\bibliographystyle{apalike}
\usepackage[pdftex]{lscape}
\usepackage[toc,page]{appendix}
% \usepackage{float}
% \usepackage{longtable}
% \usetikzlibrary{arrows}
%%%%
\title{
Supporting Information to Estimating the Initial Susceptible Fraction in Dengue Dynamics"
}           
\author{
Flavio Code\c{c}o Coelho \\
\\
\& \\
Luiz Max de Carvalho \\
}

\date{}
% Notation defs
\def \rr {$R_{t}\ $}

@@

\subsection*{A remark on prior distributions and tail behavior of the 
distribution of $R_t$}
\label{sec:tails}

In order to decide which approach to take, it may be of use analysing the variance and tail behaviour of the derived distributions for \rr. 
Consider the case of using a flat $Uniform(0, 1)$ prior for $\theta_t$.
With $a_0 = b_0 = 1$, $a_1 = a_2$ and $b_1 = b_2 + 1$.
The beta prime (inverse beta distribution) will have heavier tails compared to the conditional distribution in (\ref{eq:densityNishiura}), thus providing more conservative confidence/credibility intervals.  
As a side note, the Bayesian approach presented in Section~\ref{sec:mantel} will give similar results to those of \citet{wilson} and \citet{wilson} for $Y_{t+1}$ and $Y_t >> 1$.
Under the flat  uniform prior for $\theta_t$, the Bayesian posterior credibility interval is nearly indistinguishable from the confidence interval proposed by \citet{clopper} for $Y_{t+1}, Y_t > 20$.
Note that the $Beta(1, 1)$ uniform prior for $\theta_t$ constitutes a poor prior choice mainly because the induced distribution for \rr is only well-defined for $b_0 > 2$.
